Semantic Linearity in a Lazy Optimising Compiler

### 2025-10-07 popl26-accepted-before-rebuttal.tex

% Precisamos disto?
% %
% \begin{tabbing}
% $\ccase{e}{z~\{\ov{\rho_i \to e_i}\}} \longrightarrow^* \ccase{e'}{z~\{\ov{\rho_i \to e_i}\}}$\`where $e'$ is in WHNF and $e$ is not\\
% $\ccase{K~\ov{e}}{z~\{K~\ov{x} \to e_i\}} \longrightarrow e_i\ov{[e/x]}[K~\ov{e}/z]$\\
% $\ccase{e}{z~\{\_ \to e_i\}} \longrightarrow e_i[e/z]$\`where $e$ is in WHNF\\
% \end{tabbing}
% %
% Recall how the operational semantics encode the evaluation of case
% expressions such that,
% when a scrutinee $K~\ov{e}$ matches a constructor pattern $K~\ov{\x[\pi]}$,
% evaluation proceeds in the case alternative corresponding to the matching
% pattern, with occurrences of $\ov{x}$ being substituted by $\ov{e}$ and
% occurrences of the case binder $z$ substituted by the whole scrutinee
% $K~\ov{e}$. Constructors and lambda expressions may otherwise match match the wildcard
% pattern, resuming evaluation on the alternative body only substituting the case
% binder by the scrutinee.

%%%%???
%%%%The linear resources used to type an expression in WHNF are exactly those which
%%%%occur to the right of $\gtrdot$ in the WHNF judgement shown above
%%%%(corresponding to the resources required to typecheck the lambda body or the
%%%%constructor arguments).

%Recalling that patterns are either a wildcard that matches anything or data
%constructors parametrised by variables to bind the constructor arguments in the
%case alternative
%
%Let's consider how resources are consumed when the case exp ...
%
%Patterns data constructors with variables to bind the constructor arguments parametrised
%
%\begin{itemize}
%\item The wildcard pattern matches against anything, and is the only pattern that matches against a lambda function
%\item We can match against the data constructor and 
%\end{itemize}
%
%Considering patterns are either data constructors and variables bounding
%A function (lambda expression) can only be matched against the wildcard pattern $\_$, but 
%
%Evaluation continues in the selected branch by
%substituting, first, the pattern variables by the (possibly unevaluated)
%expressions used as arguments to the

% \todo[inline]{Case expressions are the means by which we do evaluation and
% pattern matching -- when things are scrutinized, we evaluate them (if they
% aren't evaluated -- tag is 0), and then compare the result against multiple
% alternatives}

% \todo[inline]{Item 2.
%     Pattern matching an expression that is evaluated will not consume all
%     the resources that define that computation -- because of laziness, we only
%     evaluate things to WHNF. To fully consume a value, we need to consume all
%     the linear components of that pattern.
% }

\subsubsection{Branching on WHNF-ness}

The dichotomy between evaluation (and resource usage) of a case expression
whose scrutinee is in weak head normal form and one whose scrutinee is not
leads to one of the key insights of $\lambda^\pi_\Delta$: we must \emph{branch}
on weak head normal formed-ness to accurately type case expressions.
%
When the scrutinee is already in WHNF, the resources are
unused upon evaluation and thus available in the alternatives.
%
When it is not, resources will be consumed and cannot be used in the
alternative.
%
To illustrate, consider the following case expressions:
\[
\begin{array}{ccc}
    (1)~\lambda x.~\ccase{K~x}{z~\{\_ \to x\}} &  & (2)~\lambda x.~\ccase{free~x}{z~\{\_ \to x\}}
\end{array}
\]
The first function uses $x$ linearly, but the second does not.
%
Alternatives may also use the case binder or pattern variables, referring to, respectively,
the whole scrutinee and all its used resources; or
constructor arguments and the resources used to type them.
% Using the case binder entails using all the resources required by the scrutinee, and using a pattern
% variable implies using the resources of the corresponding constructor argument.

There are three competing ways
to use the resources from a scrutinee in WHNF in a case alternative: directly, via
the case binder, or by using pattern-bound variables.
%
Recall how $\D$-bound variables encode mutual exclusion between alternative
ways of consuming resources -- it follows that case binders and pattern-bound
variables are yet another instance of $\D$-bound variables.
%
%Resources in a scrutinee that is already in WHNF are only consumed when all
%(linear) fields of the pattern are used, satisfying the definition of
%consuming resources given in Linear Haskell.
%
% We type a case expression whose scrutinee is in weak head normal form with
This suggests the following first rule for cases of scrutinees already in WHNF:
\[
\TypeCaseWHNFIntermediate
\]

First, we assert this rule is only applicable to expressions in weak head
normal form. Second, we appeal to the typing judgement for expressions in WHNF
to determine the split of resources amongst the scrutinee
sub-expressions. Finally, we type all case alternatives with the same context, using the
$\vdash_{alt}$ judgement, introducing
the case binder $z$ in the unrestricted environment as a $\D$-bound
variable whose usage environment is the linear resources $\ov{\D_i}$ used to type the
scrutinee. Those same resources $\ov{\D_i}$ are again made available in the
linear typing environment of the alternative, similarly to how the resources
used to type a $Let$ binder are still available in the continuation.
%
%The $\vdash_{alt}$ judgement is 
%annotated with the disjoint set of linear resources $\ov{\D_i}$ used
%to type the scrutinee sub-expressions,
%and the name of the case binder $z$.
%
%
Although the main idea for typing ``WHNF case expressions'' is conveyed by
this rule, the full rule of our system is slightly more involved, as will be seen after
discussing cases of scrutinees not in WHNF.

The alternative judgement $\G;\D \vdash_{alt} \rho \to e :^z_\D \s \Rightarrow
\vp$ is used to type case alternatives, encompassing three
modes that are distinguished by the kind of arrow that is used:
for alternatives of case expressions whose scrutinee is in WHNF
($\Rightarrow_\mathsf{WHNF}$), not in WHNF ($\Rightarrow_\mathsf{NWHNF}$),
and for alternatives agnostic to the WHNF-ness of the scrutinee
($\Rightarrow$), with $\Rightarrow$ also generalizing the other two.
%
% Following the $Case_\textrm{WHNF}$ rule in which we use the $\Mapsto$ alternative
% judgement, t

First, an alternative whose pattern is a constructor with $n > 0$ linear components
under the $\Rightarrow_{\mathsf{WHNF}}$ mode is typed as:
\[
\TypeAltNWHNF
\]
The rule states that, for such a match on a scrutinee already in
WHNF, we introduce the linear components of the pattern as $\D$-bound variables
with usage environment matching the linear resources required to type the
corresponding constructor argument in the scrutinee. The resources required to
type each constructor sub-expression come annotated in the judgement (as
${:}_{\ov{\D_i}}$). Unrestricted fields of the constructor are introduced as
unrestricted variables. We note that the typing environment $\D$ always
contains the resources $\ov{\D_i}$ in uses of the alternative judgement.

Second, the rule for alternatives that match on the wildcard pattern,
regardless of the WHNF status of the scrutinee (note the use of $\Rightarrow$, which
is applicable both under $\Rightarrow_{WHNF}$ and $\Rightarrow_{NWHNF}$):
\[
\TypeAltWild
\]
To type a wildcard alternative we simply type the expression with the main
judgement, ignoring all annotations. The case binder will
have already been introduced in the environment with the appropriate usage
environment by the relevant %(WHNF or not)
case expression rule.

Lastly, consider an alternative matching on a data constructor without any
linear components. The linear resources used to type a scrutinee matching such
a pattern are fully consumed during evaluation: the resulting unrestricted
constructor application can only be well-typed under an empty linear
environment.
%
Consequently, the resources to type the scrutinee which are carried over to the
case-alternative environments by the $Case_{WHNF}$ and $Case_{NWHNF}$ rules
(see the latter below), must be reactively destroyed from branches where the
pattern is unrestricted. In failing to do this, we would allow the fully-consumed
resources to be used again in the branch.
% Really, in the WHNF case it is not really necessary to remove the things
% since they must already be empty be definition.
%
%This definition agrees with the intuition we have developed by example in the
%previous section, and with the typing rule we devised for alternatives matching
%constructors without linear components.
%
%Taking into account that case expressions introduce the linear resources of the
%scrutinee in the typing environment of all alternatives, and in the usage
%environment of the case binder, we must reactively update the typing
%environments after matching on such a pattern.
%
The $Alt0$ rule essentially encodes this insight, and is applicable regardless
of the WHNF status of the scrutinee ($\Rightarrow$ mode), as long as
the constructor pattern has no linear fields:
%
\[
\TypeAltZero
\]
The rule deletes the annotated scrutinee environment $\D_s$ from
%two select environments:
the linear typing context (written $\D[\cdot/\D_s]$, a substitution of
the scrutinee typing environment by the empty linear environment $\cdot$)
and from the usage environment of the case binder $z$ (written $\G[\cdot/\D_s]_z$
to similarly delete $\D_s$ from the usage environment of the $\D$-var $z$ in $\G$).
Given that the case-binder $z$ is always annotated with $\D_s$ (essentially,
the linear environment typing the scrutinee), this latter substitution makes
the case-binder unrestricted in branches with unrestricted patterns.
%-- agreeing with our observation in Section \label{sec:semantic-linearity-examples}.

% \begin{itemize}

% \item The linear typing environment, effectively deleting the resources from
% the scrutinee made available by the case expressions (written
% $\D[\cdot/\D_s]$, a substitution of the scrutinee typing environment by the
% empty linear environment $\cdot$).

% \item The usage environment of the case binder $z$, written $\G[\cdot/\D_s]_z$
% to denote replacing the usage environment of the variable $z$ in $\G$, which
% is necessarily $\D_s$ (since we always annotate the judgement with the
% environment of the scrutinee), by the empty environment.

% \end{itemize}
%
This rule observes that any linear resources required to type an expression
that reduces to an unrestricted constructor application are fully consumed
through evaluation and the resulting unrestricted expression can be freely
discarded or duplicated (e.g. via the case binder).
%
%Reactively deleting the resources that typed the scrutinee from the linear
%environment ensures that when we match on an unrestricted pattern there are no
%remaining linear resources to consume. Otherwise, e.g., $\ccase{K_1~x}{\{K_2 \to
%K_2,K_1~y \to K_1~y\}}$ would not be well-typed since the resource $x$ would
%be left \emph{un}-consumed in the first alternative, even though we know by
%matching on $K_1$ that it has already been consumed.
%
%Furthermore, since the case binder in such an alternative refers to the
%unrestricted expression, the case binder too may be used unrestrictedly, which
%we allow by emptying its usage environment.

%It might seem as though deleting the resources from the environment in $\textsc{Alt}_0$
%is necessary to guarantee a resource is not used after it is consumed.
%%
%However, let us consider two situations -- pattern matches in a case
%expression whose scrutinee is in WHNF, and matches on a case expression whose
%scrutinee is \emph{not} in WHNF:
%%
%%\begin{enumerate}
%When the scrutinee is in WHNF, it is either an unrestricted expression
%against which any match will only introduce unrestricted variables, or an
%expression that depends on linear resources. The former trivially allows
%any resource from the scrutinee in the alternatives as well. The latter is
%further divided: 
%%\begin{enumerate}
%%
%%\item
%The pattern is unrestricted while the
%scrutinee is not, so entering this branch is impossible as long as the case
%expression is well-typed; by contradiction, the linear resources from the
%scrutinee could occur unrestrictedly in that branch.
%%
%For uniformity, we type such alternatives as those for scrutinees that
%are not in WHNF;
%%
%The pattern is linear and matches the scrutinee, in which case the
%$AltN_{\textrm{WHNF}}$ is applicable instead of $Alt0$;
%%
%The pattern is linear but does not match the scrutinee and so any
%resource could be used in such alternatives. For uniformity,
%it is also typed as though the scrutinee were not in WHNF.

%If the scrutinee is not in WHNF, the resources occurring in the
%scrutinee will be consumed when evaluation occurs. Therefore, the resources
%used in the scrutinee cannot occur in the alternative body
%(e.g. $x$ cannot occur in $e$ in $\ccase{close~x}{\{K_1 \to e\}}$)
%-- regardless of the pattern.
%%
%% In fact, the resources from a scrutinee that is not in weak head normal form
%% cannot occur in any of the alternatives, even ones matching on constructors
%% with linear components, as the resources may have been consumed when evaluating
%% the expression to weak head normal form.
%%
%We guarantee resources from a scrutinee that is not in WHNF
%cannot occur/directly be used in any case alternative in our rule for typing
%cases not in WHNF, which we introduce below.

%\end{enumerate}

\subsubsection{Proof irrelevant resources}

When the scrutinee is not in WHNF, it is not statically known which resources
will have been consumed after evaluating said scrutinee, or which
are transferred into the branch continuation, e.g. as in $\ccase{f\ x}{z\ \{K\
a \rightarrow e\}}$:
\begin{enumerate}

    \item Cases with non-WHNF scrutinees must be typed such that no resource
    from the scrutinee can be used directly in the branch again, or we
    risk
    consuming the same resource twice. In the example, $x$ must not directly occur
    in $e$ since $f$ \emph{could} have consumed it.

    \item On the other hand, since evaluation to WHNF does not guarantee all
    resources are consumed (unless the result matches an
    unrestricted pattern), we must ensure that any resources that transfer from the
    scrutinee to the alternative are ultimately consumed (in the example, $x$
    is not consumed by evaluating the scrutinee for $f = K$).

    Considering (1), we must resort to a more uniform method of guaranteeing we
    ``finish consuming'' the scrutinee. There are only two ways of
    uniformly referring to the remaining linear resources in the newly-evaluated scrutinee:

        \begin{itemize}
            \item The case binder, which refers to the whole result of evaluating the scrutinee;
            \item The linear components of a constructor pattern, which refer to expressions that may contain linear resources.
        \end{itemize}

%To guarantee all post-evaluation unconsumed scrutinee resources are consumed in
%the alternative, without statically knowing
%Statically we can't know which resources are ``transferred'', but  guarantee
%
\end{enumerate}

%is not sufficient to
%consume all resources used by the scrutinee, since sub-expressions such
%as constructor arguments will be left unevaluated and may still contain
%(``parts of'') the linear resources. To \emph{fully} consume all
%resources occurring in the scrutinee, all linear components of the
%constructor must also be fully evaluated, as witnessed by the
%$\textsc{Alt}_0$ rule above.
%%
%In essence, to type cases scrutinizing expressions not in WHNF, we need to
%forbid the usage of the scrutinee resources while tracking that we necessarily
%``finish consuming'' them, either via the case binder, or by linearly using all
%linear pattern-bound variables introduced in the alternative.

% We tackle this in due time, in the proof irrelevance section.
% \begin{itemize}

% \item The scrutinee resources must \emph{not} be used directly in the case alternatives;
% \item But the result of evaluating the scrutinee to WHNF must still be
% consumed, as all sub-expressions of the scrutinee remain unevaluated and must
% be consumed.
% \item Since the scrutinee resources cannot be consumed directly, they must be
% consumed indirectly through $\D$-variables, namely, either the case binder, or
% the linear pattern-bound variables introduced in the alternative.

% \end{itemize}

% In alternatives of a case where the scrutinee is not in
% WHNF, we must also consume the result of evaluating the scrutinee to WHNF, but the
% scrutinee resources must definitely not be available for consumption. In
% practice, the result of evaluating the scrutinee must be consumed by using
% either the case binder or all the linear components of a constructor pattern,
% except for patterns matching an unrestricted pattern, which are handled with
% the $Alt0$ rule. For WHNF scrutinees, we encode mutual exclusivity between
% consuming resources directly, with the case binder, or through linear pattern
% variables, by introducing the latter two as $\D$-bound variables.  In essence,
% for the counterpart not-WHNF scrutinees, either the case binder or linear
% pattern-bound variables \emph{must} still be used to guarantee the evaluation
% result is consumed (thus their usage environment cannot be empty), but the
% scrutinee resources cannot be used directly.

To encode  linear resources that cannot be
directly used (i.e.,~to which the $\textsc{Var}$ rule is not
applicable) we introduce \emph{proof irrelevant} resources, written as linear resources
within square brackets $[ x{:}\s]$ and lifted to contexts as $[\D]$.
Proof irrelevant resources
are linear resources in every sense, meaning they must be used
exactly once. However, since proof irrelevant resources cannot be
discarded or used directly, they have to be consumed \emph{indirectly} via
$\D$-bound variables, namely, the case binder, or, in mutual exclusion, the linear pattern-bound
variables.

Hence, to type a case expression whose scrutinee is not in WHNF, we
type the scrutinee with linear resources $\D$ and the case
alternatives by introducing the case binder with a usage environment $[\D]$,
having the same proof irrelevant linear context $[\D]$ in the typing
environment, and annotating the judgement with the proof irrelevant resources
for use in the $\Rightarrow_{\textsf{NWHNF}}$ judgement:
\[
\TypeCaseNotWHNF
\]
Note how the rule is similar to the one for scrutinees in WHNF, but
requires the resources in the case binder, typing environment, and
judgement annotation to be made ``irrelevant''.

% In practice, we can't know which resources are consumed by evaluating a given
% expression. The resources become in a limbo state -- they cannot be used
% directly because they might have been consumed, but they mustn't be considered
% as consumed, because they might not have been.  We say these resources enter a
% proof irrelevant state. They must still be tracked as though they weren't
% consumed, but they cannot be used directly to construct the program. How can we
% ensure these proof irrelevant resource variables are fully consumed? With usage
% environments -- for the case binder and for the pattern variables, and
% otherwise propagate

% \todo[inline]{The case binder and pattern variables will consume the scrutinee
% resources, be those irrelevant or relevant resources}

### 2025-10-07 popl26-accepted-before-rebuttal.tex

Finally, we recall the tentative $Case_\textrm{WHNF}$ rule presented before and
highlight its flaw: the $\G;\D \vdash_{alt} \rho \to e :^z_{\D_s} \s \Rightarrow_{\textsf{WHNF}}
\vp$ judgement is only well-defined for patterns $\rho$ matching the WHNF form
of the scrutinee, as the distribution of resources per constructor components
only makes sense for the constructor pattern matching the scrutinee.
% Essentially, if the scrutinee is in WHNF the matching alternative is easily
% determined, and must be treated with the specialized $\Mapsto$ judgement, which
% only applies to the matching constructor.
% %
Alternatives not matching the scrutinee could use resources arbitrarily as they
will never be executed. We uniformly treat non-matching alternatives
as if the scrutinee were not in WHNF. Having introduced proof irrelevant
resources, we can now present the full $Case_\textrm{WHNF}$ rule:
\[
\TypeCaseWHNF
\]
We note that it might seem unusual to specialize a rule for expressions in
WHNF, as programs scrutinizing an expression in WHNF are rarely
written by humans.
Yet, our system is designed to be suitable for optimising compilers
in which intermediate programs commonly scrutinize expressions in WHNF.
%
% Crucially, type preservation for the case-reduction substituting the
% case binder and pattern variables in the alternative is not possible
% to prove without branching on WHNF-ness, since otherwise the
% $\D$-substitution is not well-defined.  \todo[inline]{What is the
%   $\Delta$-substitution? Clarify}

\subsubsection{Splitting and tagging fragments}

%% NOTA -- Sempre a repetir a mesma coisa
% Intuitively, in case alternatives whose scrutinee is not in WHNF,
% % (and for scrutinees in WHNF which don't match the case alternative)
% the proof-irrelevant resources introduced by the case expression must be fully
% consumed, either via the case binder $z$, or by using all linear pattern-bound
% variables (for uniformity, we also treat alternatives that do not match a
% scrutinee in WHNF this way).

As opposed to scrutinees in WHNF, where the resources used to type a scrutinee
can be cleanly divided amongst the various sub-expressions of a constructor
application and, henceforth, each pattern variable, there is no direct mapping
between the resources typing a scrutinee not in WHNF and the usage
environments of pattern variables in any alternative.

%
% That is, there is no direct mapping between the usage environments of the
% linear pattern-bound variables and the resources used in the scrutinee.
% \todo{WAIT, there never was!
% Only when the K matches the scrutinee. Maybe we need a separate judgement for
% typing other constructors, which would be standalone and show up in this
% section too?}
%
We introduce \emph{tagged resources} as a means to enforce that all pattern-bound
variables for a scrutinee not in WHNF are either \emph{jointly} used to consume
all resources occurring in the environment, or not at all (instead, the case
binder may be used). Given linear resources $[\D_s]$ used to type a scrutinee,
and a pattern $K~\ov{x_\omega},\ov{y_i}$ with $i$ linear components, we assign
a usage environment $\D_i$ to each linear pattern variable where $\D_i$ is
obtained from the scrutinee environment \emph{tagged} with the constructor name
and linear-variable index $\lctag{\irr{\D_s}\!}{K_i}$. Then, $y_i{:}_{\D_i}\s$ is
introduced in $\G$, just like other $\D$-variables.
\[
  \begin{array}{cc}
    \TypeAltNNotWHNF & \TypeVarSplit
    \end{array}
\]
%
% The tag consists of a constructor name $K$ and an index $i$ identifying the
% position of the pattern variable among all bound variables in that pattern.
%
Having uniquely tagged the resources in the usage environment of each pattern
variable, we need only to express that (1) the pattern-bound $\D$-variables can
be used (i.e., the tagged resources need be available in the linear
environment in order for the $\textsc{Var}_\D$ rule to be applicable) and (2)
if a $K$-pattern-bound $\D$-variable is consumed, all remaining $K$-$\D$-variables,
standing for the remaining linear components of the same pattern, must also be
consumed.

To satisfy these two constraints, we introduce a $\textsc{Split}$ rule that
allows a linear resource $x{:}_1\s$ to be split into $n$ resources at a given
constructor $K$, where $n$ is the number of linear components of the
constructor, and each resource resulting from the split is \emph{tagged} with
$K$ and the positional index of a linear component (i.e. $x{:}_1\s$ can be
split into $\ov{\lctag{x{:}_1\s}{K_i}}^n$).
%
By assigning to each pattern variable a \emph{fragment} of the scrutinee
resources (with a tag), we require the scrutinee resources to be
$\textsc{Split}$ in order to use pattern variables at all.  Moreover, the
remaining tagged resources cannot be used directly, yet need to still be consumed
exactly once. Thus, the choice to consume the tagged resources via the usage
environments of the other pattern variables is forced, ensuring that no
variable for a constructor's linear component can go unused (the tagged environments are disjoint).
%
Tagged resources are inspired by fractional permissions in separation logic~\cite{10.5555/1760267.1760273}.

For instance, in the term $\lambda x.~\ccase{f~x}{z~\{K~a~b\to
  (a,b)\}}$, where $x$ is a linear variable, the case alternative is
typed with $\irr{x}$,
the case binder $z$ is introduced as $\z[\irr{x}]$, and the pattern variables
are introduced as $\var[a][\lctag{\irr{x}}{K_1}]$ and
$\var[b][\lctag{\irr{x}}{K_2}]$, assuming both components of $K$ are linear.
%
The occurrences of $a$ and $b$ are well-typed because we can first
$\textsc{Split}$ $[x{:}_1\s]$ into $\lctag{[x{:}_1\s]}{K_1}, \lctag{[x{:}_1\s]}{K_2}$,
%
% since for any of the scrutinee
% resources to be used by a linear pattern-bound variable usage, the
% resources must be $Split$ for the fragments corresponding to that
% $\D$-var to be consumed, and, consequently, the remaining fragments
% have to be consumed through the other linear pattern-bound variables.
%
noting that $\textsc{Split}$ can be applied both to relevant and proof irrelevant linear
resources in $\D$.

% Using tags for fragments instead of fractions (e.g. $\D_s*i/n$) is necessary
% to guarantee we cannot use the same variable multiple times to consume
% multiple fractions of the resource. It also has the added benefit of allowing
% mixing of pattern variables bound at different alternatives (e.g.~$\lambda
% x~y.~\ccase{(x,y)}{(a,b)\to\ccase{(a,b)}{(z,w)\to(a,w)}}$).

%%%%% \section{Linear Core Examples}
%%%%% 
%%%%% \todo[inline]{If I have no time...}
%%%%% 
%%%%% Linear Mini-Core~\cite{cite:minicore} lists examples of Core programs where
%%%%% semantic linearity must be understood in order for them to be well-typed. In
%%%%% this section, we show those examples in Linear Core ($\lambda^\pi_\Delta$),
%%%%% briefly explaining why they are indeed well-typed.
%%%%% 
%%%%% \paragraph{Equations}
%%%%% 
%%%%% The Linear Haskell function is compiled in Linear Core as\\
%%%%% %
%%%%% \begin{minipage}{0.47\textwidth}
%%%%% \begin{code}
%%%%% data C = Red | Green | Blue
%%%%% f :: C -o C -o C
%%%%% f Red q = q
%%%%% f p Green = p
%%%%% f Blue q = q
%%%%% \end{code}
%%%%% \end{minipage}
%%%%% \begin{minipage}{0.47\textwidth}
%%%%% \[
%%%%% \begin{array}{ll}
%%%%% \lambda p{:}_1C~q{:}_1C.~\ccase{p}{p2{:}_{\{p\}}C \\
%%%%% \{Red \to q \\
%%%%% ; \_ \to \ccase{q}{q2{:}_{\{q\}}C\\
%%%%%   \{Green \to p2\\
%%%%%   ; \_ \to \ccase{p2}{p3{:}_{\{p\}}C \\
%%%%%   \{Blue \to q2\}} \}} \}}
%%%%% \end{array}
%%%%% \]
%%%%% \end{minipage}
%%%%% 
%%%%% \paragraph{Unrestricted Fields}
%%%%% 
%%%%% The following is well-typed:
%%%%% Let $\datatype{K}{K : A \lolli B \to C}$, and $f$:
%%%%% \[
%%%%% \lambda \xl.~\ccase{x}{\var[z][x]~\{ K~a~b \to (z, b) \}}
%%%%% \]
%%%%% 
%%%%% \paragraph{Wildcard}
%%%%% 
%%%%% The following is ill-typed:
%%%%% \begin{code}
%%%%% f = \x -> case x of z { _ -> True }
%%%%% \end{code}
%%%%% 
%%%%% \paragraph{Duplication}
%%%%% 
%%%%% The following is ill-typed:
%%%%% \begin{code}
%%%%% data Foo = Foo A
%%%%% f = \x -> case x of z { Foo a -> (z, a) }
%%%%% \end{code}

% }}}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% {{{ Metatheory
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Metatheory of Linear Core\label{sec:main:metatheory}}

We develop the metatheory of Linear Core by first 
presenting the operational semantics of $\lambda^\pi_\D$,
consisting of a lazy natural semantics in the
style of Launchbury~\cite{10.1145/158511.158618}.
This semantics conveys the expected behavior of an implementation of
lazy evaluation and is agnostic to any linearity information.
Such a semantics is ill-suited for reasoning about
linearity and so we develop an \emph{instrumented} linearity-aware semantics that is
enriched with sufficient information to allow us to establish type
safety, thus showing that linearity is preserved. We show that the natural
and instrumented semantics are bisimilar for well-typed terms and so
derive type safety in the natural semantics
(Section~\ref{sec:typesafe}). Finally, we show that multiple optimising
transformations such as those performed by GHC on Linear Core are
type preserving
(Section~\ref{sec:optimisations-preserve-types-meta}), and thus
linearity preserving.

\subsection{Operational Semantics}

The Launchbury-style natural semantics, presented in Figure~\ref{fig:opsem}, captures
standard lazy evaluation while ignoring linearity information.
% The second
% semantics is significantly instrumented with type information and is thus
% \emph{linearity aware}, resulting in a stuck state when a linear
% variable is used more than once. The former is a more direct model of
% an implementation of lazy evaluation, however it is not so convenient
% to reason about formally. The latter semantics is augmented with the additional
% information necessary to strengthen the induction hypothesis to make our
% type safety proofs go through.
% %
% We show that for well-typed terms, the two semantics coincide. 
%
%Both semantics are
The natural semantics is equipped with an evaluation environment that
maps variables to expressions and is
mutated in order to express shared evaluation. Terms are explicit about
their sharing (in the sense of shared reductions in lazy evaluation),
and so we perform a translation step that makes all sharing explicit
through let-binders (see Appendix~\ref{app:opsem}).
We write $\llet{\ov{x_i{:}_{\Delta_i}\sigma = e_i}}{e'}$ for the iterated
let-binding of variables $\ov{x_i}$.

\begin{figure}

  {\small
    \[
  \begin{array}{c}
    \infer[]
    {\,}
    {\Theta : \Lambda p.e \Downarrow \Theta : \Lambda p.e}
    \qquad
    \infer[]
    {\Theta : e\Downarrow \Theta' : \Lambda p.e' \quad \Theta' :
    e'[\pi/p]\Downarrow \Theta'' : v}
    {\Theta : e~\pi\Downarrow \Theta'' : v }\\[1.5em]
    \infer[]
    {\,}
    {\Theta : \lambda x:_\pi \sigma . e \Downarrow \Theta : \lambda x:_\pi \sigma . e }
    \qquad
    \infer[]
    {\Theta : e \Downarrow \Theta' : \lambda y:_\pi \sigma . e' \quad
    \Theta' : e'[x/y] \Downarrow \Theta'' : v }
    {\Theta : e~x \Downarrow \Theta'' : v}\\[1.5em]
    \infer[]
    {(\Theta,  x:_\omega \sigma = e) : e \Downarrow \Theta' : v}
    {(\Theta , x:_\omega \sigma = e) : x \Downarrow (\Theta',x:_\omega
    \sigma = v) : v}
    \qquad
    \infer[]
    {(\Theta,x:_\omega \sigma = e) : e' \Downarrow \Theta' : v}
    {\Theta : \llet{x{:}_{\Delta}\sigma = e}{e'}\Downarrow \Theta' : v
    }\\[1.5em]
    \infer[]
    {\,}
    {\Theta : K~\ov{x_i} \Downarrow \Theta : K~\ov{x_i}}
    \qquad
    \infer[]
    {\Theta : e \Downarrow \Theta' : K~\ov{x_i} \quad
    \Theta' : e'\ov{[x_i/y_i]}[K~\ov{x_i}/z] \Downarrow \Theta'' : v }
    {\Theta :
    \ccase{e}{z{:}_{\Delta'}\sigma~\{{\dots,K~\ov{y_i} \to e', {\dots}}\}}
    \Downarrow \Theta'' : v}\\[1.5em]
    \infer[]
    {\Theta : e \Downarrow \Theta' : K~\ov{x_i} \quad
    \Theta' : e'[K~\ov{x_i}/z] \Downarrow \Theta'' : v }
    {\Theta :
    \ccase{e}{z{:}_{\Delta'}\sigma~\{\dots , \_\to
    e'}\} \Downarrow \Theta'' : v}
    \qquad
    \infer[]
    {% (\Gamma, \overline{x_i:_\omega \sigma_i = e_i}) : e_1 \Downarrow \Theta_1
    % : v_1 \dots  (\Gamma, \overline{x_i:_\omega \sigma_i = e_i}) : e_n
    % \Downarrow \Theta_n : v_n \quad
    (\Theta, \overline{x_i:_\omega \sigma_i = e_i}) : e' \Downarrow \Theta'
    : v
    }
    {\Theta : \lletrec{\overline{x_i{:}_{\Delta}\sigma_i = e_i}}{e'}
    \Downarrow \Theta' : v}
    \end{array}
\]}
    \caption{Natural Semantics of $\lambda^\pi_\D$\label{fig:opsem}}
  \end{figure}

  The natural semantics is defined via the relation
$\Theta : e \Downarrow \Theta' : v$ where $e$ is an expression, $v$ a
value, and $\Theta$ and $\Theta'$ are evaluation environments with bindings of
the form $x:_\omega\sigma = e$ which assigns the expression $e$ to the variable
$x$ of the given type $\s$. The rules are standard, augmenting the runtime
environment with let-bound expressions which are evaluated (only once) if the
corresponding let-bound variable is forced.

% The second
% semantics is significantly instrumented with type information and is thus
% \emph{linearity aware}, resulting in a stuck state when a linear
% variable is used more than once. The former is a more direct model of
% an implementation of lazy evaluation, however it is not so convenient
% to reason about formally. The latter semantics is augmented with the additional
% information necessary to strengthen the induction hypothesis to make our
% type safety proofs go through.
% %
% We show that for well-typed terms, the two semantics coincide.

While the natural semantics is a mostly direct model of an implementation
of lazy evaluation, it is inconvenient for formal reasoning about
well-typed linear terms.
%
We follow an approach
similar to that of \cite{cite:linearhaskell}, defining
an \emph{instrumented} operational semantics in which linear variables are
erased from 
the runtime environment once they are forced, ensuring that any term violating linearity
will result in a stuck state. The semantics is also type-aware and
carries with it sufficient data to reconstruct typing derivations for
the purposes of showing type safety.

The instrumented semantics is presented in
Figure~\ref{fig:linopsem}, defining the judgment $\Gamma ; \Delta \vdash (\Theta \mid e) \Downarrow (\Theta'
\mid v) : \sigma , \Sigma$ where $\Gamma$ and $\Delta$ are typing
contexts for expressions in $\Sigma$
($\Gamma$ tracks unrestricted and $\Delta$-bound variables, $\Delta$ tracks linear
assumptions);
$\Theta$ and $\Theta'$ are evaluation environments, consisting of bindings of the form
$x :_\pi \sigma = e$ or $x :_\Delta \sigma = e$; $e$ is the expression being evaluated and $v$
its resulting value, both of type $\sigma$; $\Sigma$ is a list of assignments of the form
$e : \tau$ which are expressions in which
bindings in $\Theta$ may also be used. This additional data is needed to
inductively show that the overall evaluation state is well-typed (and
linearity preserving). The rules mostly mirror those of the natural
semantics, with the exception of the linear variable rule which erases
the binding from the execution environment. Both semantics define
evaluation relations of the form $a \Downarrow b$, where $a$ and $b$
are evaluation states. We also rely on a notion
of partial derivations and partial evaluation (see
Appendix~\ref{app:opsem}), written $a\Downarrow^* b$, which allows us
to state and reason about a progress property.
